# Methods

## Data sources
(We describe our data sources, our methods in this chapter)

Storm tracks data can be downloaded from [National Hurricane Center and Central Pacific Hurricane Center](https://www.nhc.noaa.gov/data/#hurdat). The data using in the project is known as Atlantic hurricane database ([HURDAT2](https://www.nhc.noaa.gov/data/hurdat/hurdat2-format-atlantic.pdf)) 1851-2018 ([5.9MB download](https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2018-051019.txt)). The data has a comma-delimited, text format with six-hourly information on the location, maximum winds, central pressure, and (beginning in 2004) size of all known tropical cyclones and subtropical cyclones.

## Read data

```{r include=FALSE}
 # keep this chunk in your .Rmd file
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(stringr)
# Read in data set
storm_strings <- read_lines("https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2018-051019.txt")
```

## Data transformat

Describe the process of getting the data into a form in which we could work with it in R.

## Missing values

Describe any patterns we discover in missing values.
