# Methods

## Data sources
(We describe our data sources, our methods in this chapter.ion)

Storm tracks data can be downloaded from [National Hurricane Center and Central Pacific Hurricane Center]("https://www.nhc.noaa.gov/data/#hurdat"). The data using in the project is known as Atlantic hurricane database ([HURDAT2]("https://www.nhc.noaa.gov/data/hurdat/hurdat2-format-atlantic.pdf")) 1851-2018 ([5.9MB download]("https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2018-051019.txt")). The data has a comma-delimited, text format with six-hourly information on the location, maximum winds, central pressure, and (beginning in 2004) size of all known tropical cyclones and subtropical cyclones.

## read data

```{r setup, include=FALSE}
 # keep this chunk in your .Rmd file
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(stringr)
# Read in data set
storm_strings <- read_lines("https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2018-051019.txt")
```

## Data transformat

Describe the process of getting the data into a form in which you could work with it in R.

## Missing values

Describe any patterns you discover in missing values.


** Reading **

* [HURDAT]("https://cran.r-project.org/web/packages/HURDAT/HURDAT.pdf)
* [rrricanes]("https://github.com/ropensci/rrricanes")
* [Storm tracks data]("https://dplyr.tidyverse.org/reference/storms.html")
* [Visualizing Hurricane Data with Shiny]("https://www.r-bloggers.com/visualizing-hurricane-data-with-shiny/")
* [Chapter 6: Hurricane Data]("https://rpubs.com/jelsner/4826")
* [Mapping Global Earthquakes and Hurricane tracks with R]("http://david-lallemant.com/mapping-global-earthquakes-and-hurricane-tracks-with-r-2/")
* [How would you visulize hurricanes]("http://www.storytellingwithdata.com/blog/2017/10/1/how-youd-visualize-hurricanes")
* [Tracking Atlantic Hurricanes Using Statistical Methods]("https://scholarcommons.usf.edu/cgi/viewcontent.cgi?article=5927&context=etd")
