# Data Transformation {#datatrans}

<!--test-->

The ([HURDAT2](https://www.nhc.noaa.gov/data/hurdat/hurdat2-format-atlantic.pdf)) data has a comma-delimited, text format with six-hourly information on the location, maximum winds, central pressure, and (beginning in 2004) size of all known tropical cyclones and subtropical cyclones. The dataset is a combination of serveral subsets. Each subset is used for a storm track record which includes header information and values.
please refer to [this file](https://www.nhc.noaa.gov/data/hurdat/hurdat2-format-atlantic.pdf) for detail information.


```{r include=FALSE}
 # keep this chunk in your .Rmd file
 knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r echo=FALSE}
library(tidyverse)
library(stringr)
library(GGally)
# Read in data set
#dfile <- read_lines("https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2018-120319.txt")
dfile<- "../data/raw/hurdat2-1851-2018-120319.txt"
hurdat2_in <- read_lines(dfile)
```

Firstly, we extract storm `id`, `name`, and `subtext length` from each subtext header, then read in data according to each `subtext length`, and merge data subset by indexing it with storm `id`, `name`. In the original text, you will find that the `name` is non-unique. Currently, there are six lists that are used in rotation and re-cycled every six years, i.e., the 2013 list is used again in 2019. For more information, please see tropical cyclone names. To avoid the future confusion, we create `storm-id` variable by combining `name` and `year` together (e.g., _Sandy-2012_). In the original file, there are storms labeled with NAMEs but others labelled with `UNNAME`. Here, we use `name_id` variable to indicate whether a storm has a name or not. 


```{r echo=FALSE}
# read hearder information
library(stringr)
header_locations <- (1:length(hurdat2_in))[stringr::str_count(hurdat2_in, "\\,") == 3]

header_df <- readr::read_csv(hurdat2_in[header_locations], 
                             col_names = FALSE) %>% 
  dplyr::select(-c("X4"))

names(header_df) <- c("id","name","n_entries")

header_df <-  header_df %>% mutate(header_loc = as.numeric(header_locations))
```


```{r echo=FALSE}
#tail(header_df)
```


```{r echo=FALSE}
# read data value
hurdat2_df <- vector("list", nrow(header_df))
names(hurdat2_df) <- header_df$id
df_names <- c(
  "date", "time", "record_identifier", "status", "latitude", "longitude", "max_wind", "min_pressure",
  "extent_34_NE", "extent_34_SE", "extent_34_SW", "extent_34_NW",
  "extent_50_NE", "extent_50_SE", "extent_50_SW", "extent_50_NW",
  "extent_64_NE", "extent_64_SE", "extent_64_SW", "extent_64_NW", "nas"
)
```


```{r echo=FALSE}
#
for (i in seq_along(header_df$id)) {
  hurdat2_df[[i]] <- read_csv(dfile,
    skip = header_df$header_loc[i],
    n_max = header_df$n_entries[i],
    col_names = df_names,
    na = c("", "-99", "-999"),
    col_types = list(
      time = col_character(),
      record_identifier = col_character(),
      min_pressure = col_integer(),
      extent_34_NE = col_integer(),
      extent_34_SE = col_integer(),
      extent_34_SW = col_integer(),
      extent_34_NW = col_integer(),
      extent_50_NE = col_integer(),
      extent_50_SE = col_integer(),
      extent_50_SW = col_integer(),
      extent_50_NW = col_integer(),
      extent_64_NE = col_integer(),
      extent_64_SE = col_integer(),
      extent_64_SW = col_integer(),
      extent_64_NW = col_integer()
    )
  )
}
```


```{r echo=FALSE}
#hurdat2_df[[1]]
```


Secondly, we estimate `category` variable from wind speed based on [Saffir-Simpson storm category](https://www.nhc.noaa.gov/aboutsshws.php), calculate the diameter of the area experiencing hurricane strength winds (64 knots or above), `_ts_diameter_` from extent of 34 kt wind radii maximum extent in northeastern quadrant (in nautical miles, `extent_34_NE `), 34 kt wind radii maximum extent in southeastern quadrant (in nautical miles, `extent_34_SW`), 34 kt wind radii maximum extent in northeastern quadrant (in nautical miles, ` extent_34_NW `), and 34 kt wind radii maximum extent in southeastern quadrant (in nautical miles, ` extent_34_SE`), `_hu_diameter_` from extent of 64 kt wind radii maximum extent in northeastern quadrant - `extent_64_NE `, southeastern quadrant -  `extent_34_SW`, northeastern quadrant, ` extent_64_NW `), and southeastern quadrant - ` extent_64_SE`.

```{r echo=FALSE}
# Combine and clean the data sets
library(lubridate)
hurdat2 <- 
  hurdat2_df %>%
  dplyr::bind_rows(.id = "id") %>%
  dplyr::mutate(
    date = lubridate::ymd(date),
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date),
    hour = as.numeric(stringr::str_sub(time, 1, 2)),
    #datetime = as.Date(ISOdate(year, month, day, hour, min = 0, sec = 0, tz = "GMT")),
    datetime = lubridate::ymd_h(paste(year, month, day, hour, sep="-")),
    #lat_hemisphere = stringr::str_sub(latitude, -1),
    latitude = dplyr::if_else(stringr::str_sub(latitude, -1) == "N",
                              as.numeric(stringr::str_sub(latitude, 1, -2))*1, 
                              as.numeric(stringr::str_sub(latitude, 1, -2))*(-1)),
    longitude = dplyr::if_else(stringr::str_sub(longitude, -1) == "E",
                              as.numeric(stringr::str_sub(longitude, 1, -2))*1, 
                              as.numeric(stringr::str_sub(longitude, 1, -2))*(-1)),
    category = cut(max_wind, # Saffir-Simpson Hurricane Wind Scale
      breaks = c(0, 34, 64, 83, 96, 113, 137, 500),
      labels = c(-1, 0, 1, 2, 3, 4, 5),
      include.lowest = TRUE, ordered = TRUE
    ),
    # wind = wind * 1.15078, # transforms knots to mph,
    TSradius1 = extent_34_NE + extent_34_SW,
    TSradius2 = extent_34_NW + extent_34_SE,
    ts_diameter = pmax(TSradius1, TSradius2) * 1.15078, # to convert from nautical miles to miles # pmax: returns the parallel maxima and minima of the input values
    HUradius1 = extent_64_NE + extent_64_SW,
    HUradius2 = extent_64_NW + extent_64_SE,
    hu_diameter = pmax(HUradius1, HUradius2) * 1.15078, # to convert from nautical miles to miles # pmax: returns the parallel maxima and minima of the input values
    status = recode(status,
                    "TD" = "tropical depression", # maximum sustained winds below 39 mph
                    "TS" = "tropical storm",  # 39-73 mph
                    "HU" = "tropical hurricane", #74-95, 96-110, 111-130, 131-155, >156
                    "EX" = "Extratropical cyclone", ##
                    "SD" = "subtropical depression", #<18m/s or <35
                    "SS" = "subtropical storm",
                    "LO" = "a low",
                    "WV" = "tropical wave",
                    "DB" = "disturbance")
  ) 
```
<!--test: category has been calculated based on [Saffir-Simpson Hurricane Wind Scale](https://www.nhc.noaa.gov/aboutsshws.php) to indicate "Types of Damage Due to Hurricane Winds". -->

```{r echo=FALSE}
# merge header information to data values
header_df_selected <- header_df %>% dplyr::select(c("id","name"))
# headers_df_selected
hurdat2_add_name <- left_join(header_df_selected, hurdat2, by=c("id")) %>% 
  dplyr::select(id, name, datetime, year, month, day, hour, record_identifier, latitude, longitude, status, category,
         max_wind, min_pressure, ts_diameter, hu_diameter)
```
```{r echo=FALSE}
hurdat2_out <- hurdat2_add_name %>% 
  mutate(name= dplyr::if_else(grepl("UNNAMED", name), name,
                              stringr::str_to_title(name)))
hurdat2_out$status <- factor(hurdat2_out$status)
hurdat2_out$category <- factor(hurdat2_out$category)
```


```{r echo=FALSE}
# absorb header information to data values
header_df_selected <- header_df %>% dplyr::select(c("id","name"))
# headers_df_selected
hurdat2_add_name <- left_join(header_df_selected, hurdat2, by=c("id")) %>% 
  dplyr::select(id, name, datetime, year, month, day, hour, record_identifier, latitude, longitude, status, category,
         max_wind, min_pressure, ts_diameter, hu_diameter)
```

```{r echo=FALSE}
hurdat2_out <- hurdat2_add_name %>% 
  mutate(name= dplyr::if_else(grepl("UNNAMED", name), name,
                              stringr::str_to_title(name)))
hurdat2_out$status <- factor(hurdat2_out$status)
hurdat2_out$category <- factor(hurdat2_out$category)
```

```{r echo=FALSE}
#levels(hurdat2_out$status)
#hurdat2_out %>% dplyr::filter(status == "ET")
```

<!--test:*Note: there is an "ET" in _Status of system_, which does not included in the description [HURDAT2](https://www.nhc.noaa.gov/data/hurdat/hurdat2-format-atlantic.pdf). This is a typo in the dataset, `recode` it into 'EX".*-->

```{r echo=FALSE}
hurdat2_out$status <- dplyr::recode(hurdat2_out$status, ET = "Extratropical cyclone")
```
```{r echo=FALSE}
#hurdat2_out$status %>% unique()
```

```{r echo=FALSE}
hurdat2_out$category <- factor(hurdat2_out$category, levels=c("0","1","2","3","4","5"))
```

```{r echo=FALSE}
hurdat2_out$storm_id <- paste(hurdat2_out$name, hurdat2_out$year, sep="-")
#df$date <- lubridate::ymd_h(paste(df$year, df$month, df$day, df$hour, sep="-"))
```


```{r echo=FALSE}
#names(df)
```


<!--Save transformed data** for further use.-->

Thirdly, we estimate the storm duration `tc_dur_track` to those with maximum sustained surface winds of at least 35 knot and defined storms and define `tc_dur_type` for type of the duration. Here,  `S` indicates storms with duration of 2.0 days or less and will be mentioned in the following text as *short-lived* storms, and ` L` represnts storms with duration of more than 2.0 days and will be referred as “medium-to-long lived” storms.


```{r echo=FALSE}
hurdat2_out <- hurdat2_out %>% 
  #dplyr::filter(status == c("tropical hurricane", "tropical storm")) %>% 
  group_by(id) %>% mutate(dur_track=as.numeric(
  max(datetime) - min(datetime), units = "days")) %>% ungroup() %>% 
  mutate(dur_type = dplyr::if_else(dur_track<=2, "S","L"))%>% 
                                   #if_else((dur_track<=4 & dur_track>2), "M", "S")
                                   #)) %>%
  mutate(dur_type=factor(dur_type, levels = c("S", "L")))
```
```{r echo=FALSE}
#dim(hurdat2_out)
```



```{r echo=FALSE}
tcs <- c("tropical storm", "tropical hurricane","subtropical storm")
df <- hurdat2_out %>% dplyr::filter(status %in% tcs)
#df$status <- factor(df$status, levels=tcs)
df$status <- factor(df$status, levels=tcs)
```
```{r echo=FALSE}
#dim(df)
```

```{r echo=FALSE}
df <- df %>% 
  dplyr::filter(status %in% tcs) %>% 
  dplyr::group_by(id) %>% mutate(tc_dur_track=as.numeric(
  max(datetime) - min(datetime), units = "days"))%>% 
    ungroup() %>% 
  mutate(tc_dur_type = dplyr::if_else(tc_dur_track<=2, "S",
                                      "L"
                                  # if_else((tc_dur_track<=4 & tc_dur_track>2), "M", "L")
                                   )) %>%
  mutate(tc_dur_type=factor(tc_dur_type, levels = c("S", "M", "L")))
```

```{r echo=FALSE}
df$category <- (as.numeric(df$category)-1)
#df$category %>% unique()
```


```{r echo=FALSE}
df <- df %>% 
  dplyr::filter(status %in% tcs) %>% 
  dplyr::group_by(id) %>% 
  mutate(unname_label = if_else(name=="UNNAMED", "no", "yes"),
    max_category = max(category),
         max_status_label = (category == max(category)),
         max_max_wind = max(max_wind),
         min_min_pressure = min(min_pressure),
         max_ts_diameter = max(ts_diameter),
         max_hu_diamter = max(hu_diameter)) %>%
    ungroup()
#names(df)
#df$year %>% unique()
```

```{r echo=FALSE}
dir <- '../data/clean/hurricanes.csv'
readr::write_csv(df, dir)
```



Finally, the ocean surface temperature is `.nc` formate, we use `ncdf4` to read in data. Note that there is there is an "ET" typo in _Status_ of system in the [HURDAT2]( https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2018-120319.txt), which has been corrected to `EX` in the output `data\clean\hurricanes.csv` file.
<!--
_read in atmospheric CO2_
-->

```{r echo=FALSE}
d <- read.table('ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_mm_mlo.txt',
                na.strings = -99.99)
names(d) <- c('year', 'month', 'decimalYear', 'average', 'interpolated', 'trend', 'ndays')
d$time <- with(d, as.Date(paste(year, sprintf('%02d', month), '01', sep='-'), format='%Y-%m-%d', tz='UTC'))
time <- d$time
co2 <- d$interpolated
df_aco2 <- data.frame(time, co2)
#names(d)
```
```{r echo=FALSE}
#d$year %>% unique()
```

<!--
_read in HUDART (1970-2018)_
-->

```{r echo=FALSE}
hurdat_names <- c("year","Original_Named_Storms","Revised_Named_Storms",
                  "Original_Hurricanes", "Revised_Hurricanes",
                  "Original_Major_Hurricanes","Revised_Major_Hurricanes","Original_ACE","Revised_ACE",
                  "Original_US_Hurricanes","Original_US_Hurricanes_Category",
                  "Revised_US_Hurricanes", "Revised_US_Hurricanes_Category")
col_types = list(year = col_double(),
                 Original_Named_Storms = col_double(),
                 Revised_Named_Storms = col_double(),
                 Original_Hurricanes = col_double(),
                 Revised_Hurricanes = col_double(),
                 Original_Major_Hurricanes = col_double(),
                 Revised_Major_Hurricanes = col_double(),
                 Original_ACE = col_double(),
                 Revised_ACE = col_double(),
                 Original_US_Hurricanes = col_double(),
                 Original_US_Hurricanes_Category = col_character(),
                 Revised_US_Hurricanes = col_double(),
                 Revised_US_Hurricanes_Category = col_character())
dfhurdatin<- "../data/raw/HURDAT.csv"
df_hurdat <- read_csv(dfhurdatin, skip =7, n_max = 168, col_names = hurdat_names, na = c(""," "), col_types = col_types)
#names(df_hurdat)
#df_hurdat$year
```

<!--
_read in ocean temperature (1870-2010)_
-->

```{r echo=FALSE}
library(ncdf4)
 
ncfname <- "../data/raw/AMO_HADLEY.1870-2010.CLM_1901-1970.nc"
ncIn <- nc_open(ncfname)
#print(ncIn)

year <- ncvar_get(ncIn, "year")

AMO_WARM <- ncvar_get(ncIn, "AMO_WARM")
AMO_WARM_REMOVED <- ncvar_get(ncIn, "AMO_WARM_REMOVED")
AMO_WARM_REMOVED_SMTH <- ncvar_get(ncIn, "AMO_WARM_REMOVED_SMTH")
AMO_WARM_SMTH <- ncvar_get(ncIn, "AMO_WARM_SMTH")
SST_GLOBAL_MEAN <- ncvar_get(ncIn, "SST_GLOBAL_MEAN")
SST_GLOBAL_MEAN_SMTH <- ncvar_get(ncIn, "SST_GLOBAL_MEAN_SMTH")

df_amo <- data.frame(year, AMO_WARM, AMO_WARM_REMOVED, AMO_WARM_REMOVED_SMTH,AMO_WARM_SMTH,
                     SST_GLOBAL_MEAN, SST_GLOBAL_MEAN_SMTH)
#names(df_amo)
#df_amo$year
```
<!--_Save data for furture use_*-->


```{r echo=FALSE}
dir <- '../data/clean/'
write_csv(hurdat2_out, file.path(dir, "hurdat2_out.csv"))
```
```{r echo=FALSE}
#names(hurdat2_out)
```

```{r echo=FALSE}
#names(df)
```
```{r echo=FALSE}
#df$max_max_wind %>% unique()
```

**Meaning for each variables**

`_id_`

Storm id, which is unique. An id is a combination of 8 characters, 

for example, 'AL092011', 

* AL (Spaces 1 and 2) – Basin – Atlantic

* 09 (Spaces 3 and 4) – ATCF cyclone number for that year

* 2011 (Spaces 5-8, before first comma) – Year

for detail information, please see [dataformat](https://www.nhc.noaa.gov/data/hurdat/hurdat2-format-atlantic.pdf)

`_name_`

Storm Name, which is non-unique. There are six lists that are used in rotation and re-cycled every six years, i.e., the 2013 list is used again in 2019. For more information, please see [tropical cyclone names](https://www.nhc.noaa.gov/aboutnames.shtml).

`_storm_id_`

Storm name and id combined,  i.e., Sandy-2012  

`_unname_label_`   

Storms have name or not (“yes”, “no”)

`_datetime, year, month, day, hour_`

Date of report (in Universal Time Coordinate)

`_record_identifier_`

C – Closest approach to a coast, not followed by a landfall

G – Genesis

I – An intensity peak in terms of both pressure and wind

L – Landfall (center of system crossing a coastline)

P – Minimum in central pressure

R – Provides additional detail on the intensity of the cyclone when rapid changes are underway

S – Change of status of the system

T – Provides additional detail on the track (position) of the cyclone

W – Maximum sustained wind speed

`_latitude,longitude_`

Location of storm center

`_status_`

Storm classification (Tropical Depression, Tropical Storm, or Hurricane)

TD – Tropical cyclone of tropical depression intensity (< 34 knots)

TS – Tropical cyclone of tropical storm intensity (34-63 knots)

HU – Tropical cyclone of hurricane intensity (> 64 knots)

EX – Extratropical cyclone (of any intensity)

SD – Subtropical cyclone of subtropical depression intensity (< 34 knots)

SS – Subtropical cyclone of subtropical storm intensity (> 34 knots)

LO – A low that is neither a tropical cyclone, a subtropical cyclone, nor an extratropical cyclone (of any intensity)

WV – Tropical Wave (of any intensity)

DB – Disturbance (of any intensity) 

`_category_`

[Saffir-Simpson storm category](https://www.nhc.noaa.gov/aboutsshws.php) (estimated from wind speed. -1 = Tropical Depression, 0 = Tropical Storm)

`_max_wind_`

storm's maximum sustained wind speed (in knots)

`_min_pressure_`

Air pressure at the storm's center (in millibars)

`_ts_diameter_`

Diameter of the area experiencing tropical storm strength winds (34 knots or above)

`_hu_diameter_`

Diameter of the area experiencing hurricane strength winds (64 knots or above)

`_max_category_`

Maximum category of each storm track

`_max_status_label_`

Label (“TRUE”, “FALSE”) to indicate whether the measurement is for the maximum status of each track

`_max_max_wind_`  

The maximum value of the ` max_wind ` for each track

`_min_min_pressure_` 

The minimum value of the ` min_pressure ` for each track

`_max_ts_diameter_`

The maximum value of the ` ts_diameter ` for each track

`_max_hu_diamter_`

The maximum value of the ` hu_diameter ` for each track

!--test-->

```{r echo=FALSE}
d <- read.table('ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_mm_mlo.txt',
                na.strings = -99.99)
names(d) <- c('year', 'month', 'decimalYear', 'average', 'interpolated', 'trend', 'ndays')
d$time <- with(d, as.Date(paste(year, sprintf('%02d', month), '01', sep='-'), format='%Y-%m-%d', tz='UTC'))
time <- d$time
co2 <- d$interpolated
df_aco2 <- data.frame(time, co2)
#names(d)
```
```{r echo=FALSE}
#d$year %>% unique()
```



```{r echo=FALSE}
hurdat_names <- c("year","Original_Named_Storms","Revised_Named_Storms",
                  "Original_Hurricanes", "Revised_Hurricanes",
                  "Original_Major_Hurricanes","Revised_Major_Hurricanes","Original_ACE","Revised_ACE",
                  "Original_US_Hurricanes","Original_US_Hurricanes_Category",
                  "Revised_US_Hurricanes", "Revised_US_Hurricanes_Category")
col_types = list(year = col_double(),
                 Original_Named_Storms = col_double(),
                 Revised_Named_Storms = col_double(),
                 Original_Hurricanes = col_double(),
                 Revised_Hurricanes = col_double(),
                 Original_Major_Hurricanes = col_double(),
                 Revised_Major_Hurricanes = col_double(),
                 Original_ACE = col_double(),
                 Revised_ACE = col_double(),
                 Original_US_Hurricanes = col_double(),
                 Original_US_Hurricanes_Category = col_character(),
                 Revised_US_Hurricanes = col_double(),
                 Revised_US_Hurricanes_Category = col_character())
dfhurdatin<- "../data/raw/HURDAT.csv"
df_hurdat <- read_csv(dfhurdatin, skip =7, n_max = 168, col_names = hurdat_names, na = c(""," "), col_types = col_types)
#names(df_hurdat)
#df_hurdat$year
```


```{r echo=FALSE}
library(ncdf4)
 
ncfname <- "../data/raw/AMO_HADLEY.1870-2010.CLM_1901-1970.nc"
ncIn <- nc_open(ncfname)
#print(ncIn)

year <- ncvar_get(ncIn, "year")

AMO_WARM <- ncvar_get(ncIn, "AMO_WARM")
AMO_WARM_REMOVED <- ncvar_get(ncIn, "AMO_WARM_REMOVED")
AMO_WARM_REMOVED_SMTH <- ncvar_get(ncIn, "AMO_WARM_REMOVED_SMTH")
AMO_WARM_SMTH <- ncvar_get(ncIn, "AMO_WARM_SMTH")
SST_GLOBAL_MEAN <- ncvar_get(ncIn, "SST_GLOBAL_MEAN")
SST_GLOBAL_MEAN_SMTH <- ncvar_get(ncIn, "SST_GLOBAL_MEAN_SMTH")

df_amo <- data.frame(year, AMO_WARM, AMO_WARM_REMOVED, AMO_WARM_REMOVED_SMTH,AMO_WARM_SMTH,
                     SST_GLOBAL_MEAN, SST_GLOBAL_MEAN_SMTH)
#names(df_amo)
#df_amo$year
```
<!--_Save data for furture use_*-->


```{r echo=FALSE}
dir <- '../data/clean/'
write_csv(hurdat2_out, file.path(dir, "hurdat2_out.csv"))
```
```{r echo=FALSE}
#names(hurdat2_out)
```








